Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2022-01-18 manuscript_BMB.tex
manuscript back to initial submission, the man comments added to the response letter for now!

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}). Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as the members of a university), \cite{bergstrom2020frequency} provided analytical results quantifying the effects that proactive screening of asymptomatic individuals and isolation of confirmed-positive cases could have in reducing the spread of disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation; they similarly suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
\cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends almost primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from these previous efforts in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.

>>>>

\newcommand{\comment}{\showcomment}
\newcommand{\showcomment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\nocomment}[3]{}

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
% \newcommand{\rev}[1]{\comment{red}{Revised}{#1}}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

Although testing and tracing is a key part of infection control strategies, mathematical epidemiologists have typically analyzed it with detailed models designed to inform particular public health efforts \citep{endo2020implication,hellewell2020feasibility,jenness2020modeling}, rather than analyzing simple but general models of the feedback between testing and transmission dynamics.
There have been several modeling studies of testing and tracing dynamics and their interaction with epidemiological dynamics. In the context of repeated screening and random testing of isolated populations (such as the members of a university), \cite{bergstrom2020frequency} provided analytical results quantifying the effects that proactive screening of asymptomatic individuals and isolation of confirmed-positive cases could have in reducing the spread of disease. 
\cite{rogers2021high} simulated a SEIR model with testing and isolation; they similarly suggest a strategy of rapid testing with antigen tests and the subsequent isolation of confirmed-positive individuals.
% reviewer1.com1
\cite{friston2021testing} model the effects of self-isolation on testing and tracing with a focus on projections under different testing and tracing scenarios. They conclude that the emergence of a second wave depends almost primarily on the rate at which immunity is lost and that it is necessary to track asymptomatic individuals in order to control the outbreak. 
Our modeling approach differs from these previous efforts in that it examines the effects of test-return rates and of different levels of testing focus, from random to highly targeted. We hope this paper will inspire further explorations of the fundamental properties of epidemic models that incorporate explicit testing processes.

### 2022-01-18 response1_BMB.tex
manuscript back to initial submission, the man comments added to the response letter for now!

\usepackage{amssymb,amsmath,amsthm}
\usepackage{xcolor}
\usepackage{xspace}

@@

\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

\newcommand{\fady}[1]{\comment{cyan}{Fady}{#1}}
\newcommand{\ali}[1]{\comment{magenta}{Ali}{#1}}
\newcommand{\jd}[1]{\comment{blue}{JD}{#1}}
\newcommand{\david}[1]{\comment{orange}{DJDE}{#1}}
\newcommand{\bmb}[1]{\comment{red}{BMB}{#1}}
\newcommand{\todo}[1]{\comment{red}{TODO}{#1}}
\newcommand{\new}[1]{\comment{red}{NEW}{#1}}
\newcommand{\needref}{\textcolor{red}{(REFS?)}}
\newcommand{\com}[1]{\comment{red}{Comment}{#1}}
\newcommand{\res}[1]{\comment{red}{Response}{#1}}

@@

\com 1 {\it The main surprising results about the possibility of R0 increasing with increased testing intensity or decreasing with increased delays seem particularly intertwined with the behavior displayed by individuals who are waiting for their test results. In particular, having long delays can make versions of the model with high $\theta_w$ effectively treat these waiting individuals into a partially socially-distanced or quarantined compartment, while, in versions in which waiting individuals do not change their behavior, increased asymptomatic testing can provide extra time to miss new infectious. This connection between behavior of waiting and confirmed individuals and how testing impacts R0 makes me curious about a few aspects of combining testing with social distancing.

@@

\res 1 
We now have added several lines and the suggested reference [1] to the end of the discussion to acknowledge the "gateway testing" and the potential benefit of our modelling results.

@@

\res 2
[not sure here!] 
Regarding the incorporation of an exposed class or asymptomatic and infectious compartment to the current model; 
Incorporating the extra compartments (exposed, asymptomatic and infectious) would effect the testing weights and isolation efficacy parameters. In particular, we know that people know something about their exposure (assuming we don't know about symptoms). We know that there is a bias towards testing infected people, due to having some knowledge of risk. Testing people of known exposures will increase testing weight $w_I$, this is testing and tracing. Also, exposed people presumably practice more isolation, thus $\theta\_w$ will increase and consequently $\Rnum$ will decrease.

{\it While I think that the analysis already presented in the paper is sufficient for the current paper, I think that ideas related to these questions might be interesting for future research or as points for discussion.}

\com 3 
{\it Due to its important effects on R0, I think it could be useful to describe the rate $\omega$ for return of test results directly in the text (in addition to the table of parameters) before it appears in Equation 4 in the expression for the disease-free equilibrium.}

\res 3
We now have added a line in the manuscript to describe the parameters involved in the DFE before Equation (4)

@@

\res 1
We now have added a line in the manuscript (the first paragraph of the Discussion section) to clarify what we mean by the term ``caricature''. 
[ali: note sure what we should discuss more than what we have in the manuscript!]

@@

\res 2

[This can be added to the manuscript?]
The potential future work of this project include:
\begin{enumerate}
\item
Incorporating the asymptomatic and symptomatic compartments to the model and verify the present results. 
\item
Quantify the testing weights, when we incorporate asymptomatic and symptomatic compartments to the model. In particular, We know that not all symptomatic people are the same. people know something about their exposure, and not necessarily about symptoms. We know that there is a bias towards testing infected people, due to having some knowledge of risk. It is harder to quantify that than quantifying symptom/no symptom but we can say that would shift if we have a policy of screening people of high risk areas will push things towards infected people. Also testing people of known exposures will increase testing weight for infected people $w_I$, this is exactly testing and tracing. It is much harder to quantify that. In principle if we knew enough of the case report, hospitalizations, test positivity, death, we could make an estimate of $w_X$'s. Also note that $w_X$'s don't stay the same because it is all bound up with capacity because we tend to bias towards testing people who are more likely be infected and the more testing capacity we have the more we are willing to test more people less likely to be infected. Thus there is a relationship between testing capacity and the $w_X$'s because our testing policy changes as the testing capacity changes. Except for the symptomatic example, it is very hard to quantify the effect of a particular testing policy on the $w_X$'s. In short, "test positivity tells a lot about testing weights"; It is hard to know exactly how much information there is and how much we can estimate that. 
\end{enumerate}

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{enumerate}
\usepackage{natbib}

@@

\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

\newcommand{\fady}[1]{\comment{cyan}{Fady}{#1}}
\newcommand{\ali}[1]{\comment{magenta}{Ali}{#1}}
\newcommand{\jd}[1]{\comment{blue}{JD}{#1}}
\newcommand{\david}[1]{\comment{orange}{DJDE}{#1}}
\newcommand{\bmb}[1]{\comment{red}{BMB}{#1}}
\newcommand{\todo}[1]{\comment{red}{TODO}{#1}}
\newcommand{\new}[1]{\comment{red}{NEW}{#1}}
\newcommand{\needref}{\textcolor{red}{(REFS?)}}
\newcommand{\com}[1]{\comment{red}{Comment}{#1}} %reviewer's comment
\newcommand{\res}[1]{\comment{red}{Response}{#1}} %our response
\newcommand{\rev}[1]{\comment{red}{In the manuscript}{#1}} %manuscript edition

@@

\com 1 {\it The main surprising results about the possibility of R0 increasing with increased testing intensity or decreasing with increased delays seem particularly intertwined with the behavior displayed by individuals who are waiting for their test results. In particular, having long delays can make versions of the model with high $\theta\_w$ effectively treat these waiting individuals into a partially socially-distanced or quarantined compartment, while, in versions in which waiting individuals do not change their behavior, increased asymptomatic testing can provide extra time to miss new infectious. This connection between behavior of waiting and confirmed individuals and how testing impacts R0 makes me curious about a few aspects of combining testing with social distancing.

@@

\res 1 
We now have added several lines and the suggested reference [1] to the end of the discussion to acknowledge the "gateway testing" and the potential benefit of our modelling results in controling the disease spread.

\rev {\cite{muller2021mathematical} simulated a structured SEIR-type model. Their model incorporated different testing scenarios -- including gateway testing, surveillance testing, and contact tracing -- as well as individual level control measures such as mask wearing and social distancing. They indicate the benefit of gateway testing in conjunction with effective contact tracing or robust surveillance testing in controling the spread of the infection. Under such approaches, individuals must practice increased levels of social distancing or self- isolation until they have achieved a series of confirmed negative tests. It is notable that while this type of time- dependent testing and self-isolation protocol might not fit well within the framework of linear stability of a disease-free equilibrium, the relationship between increased $\theta_w$ and the benefits of a longer waiting period $1/\omega$ seems to indicate the possibility that combining waiting for results and self-isolation can help to stem early outbreaks (e.g. at the beginning of a university term).}

@@

\res 2
[not sure here!]

\rev{ Regarding the incorporation of an exposed class or asymptomatic and infectious compartment to the current model; 
Incorporating the extra compartments (exposed, asymptomatic and infectious) would effect the testing weights and isolation efficacy parameters. In particular, we know that people know something about their exposure (assuming we don't know about symptoms). We know that there is a bias towards testing infected people, due to having some knowledge of risk. Testing people of known exposures will increase testing weight $w_I$, this is testing and tracing. Also, exposed people presumably practice more isolation, thus $\theta\_w$ will increase and consequently $\Rnum$ will decrease. }

{\it While I think that the analysis already presented in the paper is sufficient for the current paper, I think that ideas related to these questions might be interesting for future research or as points for discussion.}
\com 3 
{\it Due to its important effects on R0, I think it could be useful to describe the rate $\omega$ for return of test results directly in the text (in addition to the table of parameters) before it appears in Equation 4 in the expression for the disease-free equilibrium.}

\res 3
We now have added a line in the manuscript to describe the parameters involved in the DFE before Equation (4).

\rev {The DFE depends on \percap testing intensity $\rho$, the rate of test return $\omega$, and the population size $N$.}

@@

\res 1
We now have added a line in the manuscript (the first paragraph of the Discussion section) to clarify what we mean by the term ``caricature''.

Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); \rev{the array consisting of multidirectional subtle interplay between epidemiology, behaviors and public health policies.} Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

\ali {note sure what we should discuss more than what we have in the manuscript!}

@@

\res 2
We have added some lines to the discussion section of the manuscript to include some potential future work.

\rev{ The potential future work of this project include:
\begin{enumerate}
\item
Incorporating the asymptomatic and symptomatic compartments to the model and verify the present results. 
\item
Quantify the testing weights, when we incorporate asymptomatic and symptomatic compartments to the model. In particular, We know that not all symptomatic people are the same. people know something about their exposure, and not necessarily about symptoms. We know that there is a bias towards testing infected people, due to having some knowledge of risk. It is harder to quantify that than quantifying symptom/no symptom but we can say that would shift if we have a policy of screening people of high risk areas will push things towards infected people. Also testing people of known exposures will increase testing weight for infected people $w_I$, this is exactly testing and tracing. It is much harder to quantify that. In principle if we knew enough of the case report, hospitalizations, test positivity, death, we could make an estimate of $w_X$'s. Also note that $w_X$'s don't stay the same because it is all bound up with capacity because we tend to bias towards testing people who are more likely be infected and the more testing capacity we have the more we are willing to test more people less likely to be infected. Thus there is a relationship between testing capacity and the $w_X$'s because our testing policy changes as the testing capacity changes. Except for the symptomatic example, it is very hard to quantify the effect of a particular testing policy on the $w_X$'s. In short, "test positivity tells a lot about testing weights"; It is hard to know exactly how much information there is and how much we can estimate that. 
\end{enumerate}
}
